---
title: "Homework 4 - Football Analytics"
author: "Zach Perkins"
format: 
  html:
    toc: true
    theme: cosmo
execute:
  output: false
  warning: false
  message: false
---



# Question 1

## Data Import & Preprocessing

```{r}
# Install required packages if missing:
# install.packages(c("cfbfastR", "dplyr", "tidyr", "purrr", "knitr"))

library(cfbfastR)
library(dplyr)
library(tidyr)
library(purrr)
library(knitr)
```

```{r}
seasons <- 2021:2025

raw_games <- load_cfb_schedules(seasons)
```

```{r}
games_clean <- raw_games %>%
  filter(
    completed == TRUE,
    !is.na(home_points),
    !is.na(away_points),
    home_division == "fbs",
    away_division == "fbs",
    season_type %in% c("regular")
  ) %>%
  select(
    season,
    week,
    home_team,
    away_team,
    home_points,
    away_points
  ) %>%
  arrange(season, week)
```


## SRS Model

```{r}
calc_srs_scratch <- function(df) {
  teams <- unique(c(df$home_team, df$away_team))
  n_games <- nrow(df)
  
  # Construct design matrix: Home (+1), Away (-1)
  X <- matrix(0, nrow = n_games, ncol = length(teams))
  colnames(X) <- teams
  
  for (i in seq_len(n_games)) {
    X[i, df$home_team[i]] <- 1
    X[i, df$away_team[i]] <- -1
  }
  
  margin <- df$home_points - df$away_points
  
  # Intercept represents the home-team effect
  fit <- lm(margin ~ 1 + X)
  
  coefs <- coef(fit)
  
  # Save home-team effect separately
  home_effect <- unname(coefs["(Intercept)"])
  
  # Keep only team coefficients
  srs_vec <- coefs[-1]
  names(srs_vec) <- gsub("^X", "", names(srs_vec))
  srs_vec[is.na(srs_vec)] <- 0
  
  # Center team ratings around zero
  srs_centered <- srs_vec - mean(srs_vec, na.rm = TRUE)
  
  data.frame(
    team = names(srs_centered),
    season = df$season[1],
    SRS = round(unname(srs_centered), 2),
    home_effect = round(home_effect, 2),
    stringsAsFactors = FALSE
  )
}
```

```{r}
srs_by_season <- games_clean %>%
  group_split(season) %>%
  map(calc_srs_scratch) %>%
  bind_rows()
```


### Top and Bottom 10 SRS Teams by Season

```{r}
for (yr in sort(unique(srs_by_season$season))) {
  cat("\n### Season:", yr, "\n\n")
  
  top_10 <- srs_by_season %>%
    filter(season == yr) %>%
    arrange(desc(SRS)) %>%
    slice_head(n = 10) %>%
    mutate(Rank = row_number()) %>%
    select(
      Rank,
      Team = team,
      `SRS Rating` = SRS
    )
  
  bottom_10 <- srs_by_season %>%
    filter(season == yr) %>%
    arrange(SRS) %>%
    slice_head(n = 10) %>%
    mutate(Rank = row_number()) %>%
    select(
      Rank,
      Team = team,
      `SRS Rating` = SRS
    )
  
  cat("\nTop 10\n\n")
  print(
    kable(
      top_10,
      format = "markdown",
      align = c("c", "l", "r")
    )
  )
  
  cat("\n\nBottom 10\n\n")
  print(
    kable(
      bottom_10,
      format = "markdown",
      align = c("c", "l", "r")
    )
  )
  
  cat("\n")
}
```


## Opponent-Adjusted Game Control

```{r}
# Load college football play-by-play data
pbp_raw <- load_cfb_pbp(seasons = 2021:2025)
```


```{r}
# Filter play-by-play data and create full-game time remaining
pbp_clean <- pbp_raw %>%
  filter(
    period %in% 1:4,
    home_team_division == "fbs",
    away_team_division == "fbs",
    !is.na(clock_minutes),
    !is.na(clock_seconds),
    clock_minutes >= 0,
    clock_minutes <= 15,
    clock_seconds >= 0,
    clock_seconds <= 59,
    !is.na(score_diff),
    !is.na(pos_team),
    !is.na(home_team),
    !is.na(away_team)
  ) %>%
  mutate(
    # Convert quarter clock to regulation game seconds remaining
    game_seconds_remaining =
      (4 - period) * 900 +
      (clock_minutes * 60) +
      clock_seconds,
    
    # Convert score differential to the home team's perspective
    home_lead = ifelse(
      pos_team == home_team,
      score_diff,
      -score_diff
    )
  ) %>%
  arrange(season, game_id, desc(game_seconds_remaining)) %>%
  group_by(season, game_id) %>%
  mutate(
    play_duration =
      lag(game_seconds_remaining, default = 3600) -
      game_seconds_remaining,
    
    play_duration = ifelse(
      play_duration < 0,
      0,
      play_duration
    )
  ) %>%
  ungroup()
```


### Game Control Calculation

```{r}
# Calculate time-weighted score differential for each game
game_control_by_game <- pbp_clean %>%
  group_by(
    season,
    game_id,
    home_team,
    away_team
  ) %>%
  summarize(
    total_time = sum(play_duration, na.rm = TRUE),
    
    home_gc =
      sum(home_lead * play_duration, na.rm = TRUE) /
      ifelse(total_time > 0, total_time, 1),
    
    away_gc = -home_gc,
    
    .groups = "drop"
  )
```


```{r}
# Remove games with incomplete play-by-play coverage
game_control_by_game <- game_control_by_game %>%
  filter(total_time >= 3000)
```


### Adjusted Game Control

```{r}
calc_adj_gc_scratch <- function(df) {
  teams <- unique(c(df$home_team, df$away_team))
  n_games <- nrow(df)
  
  # Construct design matrix: Home (+1), Away (-1)
  X <- matrix(0, nrow = n_games, ncol = length(teams))
  colnames(X) <- teams
  
  for (i in seq_len(n_games)) {
    X[i, df$home_team[i]] <- 1
    X[i, df$away_team[i]] <- -1
  }
  
  # Home game control is the response variable
  margin <- df$home_gc
  
  # Intercept represents the home-team effect
  fit <- lm(margin ~ 1 + X)
  
  coefs <- coef(fit)
  
  # Save home-team effect separately
  home_effect <- unname(coefs["(Intercept)"])
  
  # Keep only team coefficients
  gc_vec <- coefs[-1]
  names(gc_vec) <- gsub("^X", "", names(gc_vec))
  gc_vec[is.na(gc_vec)] <- 0
  
  # Center team ratings around zero
  gc_centered <- gc_vec - mean(gc_vec, na.rm = TRUE)
  
  data.frame(
    team = names(gc_centered),
    season = df$season[1],
    Adj_GC = round(unname(gc_centered), 2),
    home_effect = round(home_effect, 2),
    stringsAsFactors = FALSE
  )
}
```


```{r}
adj_gc_by_season <- game_control_by_game %>%
  group_split(season) %>%
  map(calc_adj_gc_scratch) %>%
  bind_rows()
```


### Top and Bottom 10 Adjusted Game Control Teams by Season

```{r}
for (yr in sort(unique(adj_gc_by_season$season))) {
  cat("\n### Season:", yr, "\n\n")
  
  top_10 <- adj_gc_by_season %>%
    filter(season == yr) %>%
    arrange(desc(Adj_GC)) %>%
    slice_head(n = 10) %>%
    mutate(Rank = row_number()) %>%
    select(
      Rank,
      Team = team,
      `Adjusted Game Control` = Adj_GC
    )
  
  bottom_10 <- adj_gc_by_season %>%
    filter(season == yr) %>%
    arrange(Adj_GC) %>%
    slice_head(n = 10) %>%
    mutate(Rank = row_number()) %>%
    select(
      Rank,
      Team = team,
      `Adjusted Game Control` = Adj_GC
    )
  
  cat("\nTop 10\n\n")
  print(
    kable(
      top_10,
      format = "markdown",
      align = c("c", "l", "r")
    )
  )
  
  cat("\n\nBottom 10\n\n")
  print(
    kable(
      bottom_10,
      format = "markdown",
      align = c("c", "l", "r")
    )
  )
  
  cat("\n")
}
```


### Average Win Probability Calculation

```{r}
# Calculate average win probability for each team in each game
awp_by_game <- pbp_raw %>%
  filter(
    period %in% 1:4,
    home_team_division == "fbs",
    away_team_division == "fbs",
    !is.na(home_wp_before),
    !is.na(away_wp_before),
    !is.na(home_team),
    !is.na(away_team)
  ) %>%
  group_by(
    season,
    game_id,
    home_team,
    away_team
  ) %>%
  summarize(
    home_awp = mean(home_wp_before, na.rm = TRUE),
    away_awp = mean(away_wp_before, na.rm = TRUE),
    .groups = "drop"
  )
```


### Season Average Win Probability

```{r}
# Reshape game-level AWP into one row per team per game
awp_team_games <- bind_rows(
  awp_by_game %>%
    select(
      season,
      game_id,
      team = home_team,
      awp = home_awp
    ),
  
  awp_by_game %>%
    select(
      season,
      game_id,
      team = away_team,
      awp = away_awp
    )
)

# Calculate unadjusted average win probability by season
season_awp <- awp_team_games %>%
  group_by(season, team) %>%
  summarize(
    games_played = n(),
    avg_wp = mean(awp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(season, desc(avg_wp))
```

### Adjusted Average Win Probability

```{r}
calc_adj_awp_scratch <- function(df) {
  teams <- unique(c(df$home_team, df$away_team))
  n_games <- nrow(df)
  
  # Construct design matrix: Home (+1), Away (-1)
  X <- matrix(0, nrow = n_games, ncol = length(teams))
  colnames(X) <- teams
  
  for (i in seq_len(n_games)) {
    X[i, df$home_team[i]] <- 1
    X[i, df$away_team[i]] <- -1
  }
  
  # Home team's average win probability is the response
  awp <- df$home_awp
  
  # Intercept represents the home-team effect
  fit <- lm(awp ~ 1 + X)
  
  coefs <- coef(fit)
  
  # Save home-team effect separately
  home_effect <- unname(coefs["(Intercept)"])
  
  # Keep only team coefficients
  awp_vec <- coefs[-1]
  names(awp_vec) <- gsub("^X", "", names(awp_vec))
  awp_vec[is.na(awp_vec)] <- 0
  
  # Center team ratings around zero
  awp_centered <- awp_vec - mean(awp_vec, na.rm = TRUE)
  
  data.frame(
    team = names(awp_centered),
    season = df$season[1],
    Adj_AWP = round(unname(awp_centered), 4),
    home_effect = round(home_effect, 4),
    stringsAsFactors = FALSE
  )
}
```

```{r}
adj_awp_by_season <- awp_by_game %>%
  group_split(season) %>%
  map(calc_adj_awp_scratch) %>%
  bind_rows()
```


### Top and Bottom 10 Adjusted AWP Teams by Season

```{r}
for (yr in sort(unique(adj_awp_by_season$season))) {
  cat("\n### Season:", yr, "\n\n")
  
  top_10 <- adj_awp_by_season %>%
    filter(season == yr) %>%
    arrange(desc(Adj_AWP)) %>%
    slice_head(n = 10) %>%
    mutate(Rank = row_number()) %>%
    select(
      Rank,
      Team = team,
      `Adjusted AWP` = Adj_AWP
    )
  
  bottom_10 <- adj_awp_by_season %>%
    filter(season == yr) %>%
    arrange(Adj_AWP) %>%
    slice_head(n = 10) %>%
    mutate(Rank = row_number()) %>%
    select(
      Rank,
      Team = team,
      `Adjusted AWP` = Adj_AWP
    )
  
  cat("\nTop 10\n\n")
  
  print(
    kable(
      top_10,
      format = "markdown",
      align = c("c", "l", "r")
    )
  )
  
  cat("\n\nBottom 10\n\n")
  
  print(
    kable(
      bottom_10,
      format = "markdown",
      align = c("c", "l", "r")
    )
  )
  
  cat("\n")
}
```


### Season Average Game Control

```{r}
# Reshape game-level Game Control into one row per team per game
gc_team_games <- bind_rows(
  game_control_by_game %>%
    select(
      season,
      game_id,
      team = home_team,
      game_control = home_gc
    ),
  
  game_control_by_game %>%
    select(
      season,
      game_id,
      team = away_team,
      game_control = away_gc
    )
)

# Calculate unadjusted average Game Control by season
season_game_control <- gc_team_games %>%
  group_by(season, team) %>%
  summarize(
    games_played = n(),
    avg_game_control = mean(game_control, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(season, desc(avg_game_control))
```

