---
title: "Homework 4 - Question 2"
author: "Zach Perkins"
format:
  html:
    toc: true
    theme: cosmo
execute:
  output: false
  warning: false
  message: false
---

# Question 2 - Out-of-Sample Predictive Performance

## Data Import & Preprocessing

```{r}
# Install required packages if missing:
# install.packages(c("cfbfastR", "dplyr", "tidyr", "purrr", "knitr"))

library(cfbfastR)
library(dplyr)
library(tidyr)
library(purrr)
library(knitr)

seasons <- 2021:2025

raw_games <- load_cfb_schedules(seasons)

games_clean <- raw_games %>%
  filter(
    completed == TRUE,
    !is.na(home_points),
    !is.na(away_points),
    home_division == "fbs",
    away_division == "fbs",
    season_type %in% c("regular")
  ) %>%
  select(
    season,
    week,
    home_team,
    away_team,
    home_points,
    away_points
  ) %>%
  arrange(season, week)
```

```{r}
train_games <- games_clean %>%
  filter(week <= 10)

test_games <- games_clean %>%
  filter(week > 10)
```


## SRS Training Model

```{r}
calc_srs_train <- function(df) {
  teams <- unique(c(df$home_team, df$away_team))
  n_games <- nrow(df)
  
  # Construct design matrix: Home (+1), Away (-1)
  X <- matrix(0, nrow = n_games, ncol = length(teams))
  colnames(X) <- teams
  
  for (i in seq_len(n_games)) {
    X[i, df$home_team[i]] <- 1
    X[i, df$away_team[i]] <- -1
  }
  
  margin <- df$home_points - df$away_points
  
  # Include a home-team effect
  fit <- lm(margin ~ 1 + X)
  
  coefs <- coef(fit)
  
  home_effect <- unname(coefs["(Intercept)"])
  
  srs_vec <- coefs[-1]
  names(srs_vec) <- gsub("^X", "", names(srs_vec))
  srs_vec[is.na(srs_vec)] <- 0
  
  srs_centered <- srs_vec - mean(srs_vec, na.rm = TRUE)
  
  data.frame(
    team = names(srs_centered),
    season = df$season[1],
    SRS = unname(srs_centered),
    home_effect = home_effect,
    stringsAsFactors = FALSE
  )
}
```

```{r}
srs_train_ratings <- train_games %>%
  group_split(season) %>%
  map(calc_srs_train) %>%
  bind_rows()
```

### SRS Test Predictions

```{r}
srs_test_predictions <- test_games %>%
  left_join(
    srs_train_ratings %>%
      select(
        season,
        home_team = team,
        home_srs = SRS,
        home_effect
      ),
    by = c("season", "home_team")
  ) %>%
  left_join(
    srs_train_ratings %>%
      select(
        season,
        away_team = team,
        away_srs = SRS
      ),
    by = c("season", "away_team")
  ) %>%
  mutate(
    predicted_margin = home_effect + home_srs - away_srs,
    actual_margin = home_points - away_points
  )
```


### SRS Out-of-Sample Accuracy

```{r}
srs_accuracy <- srs_test_predictions %>%
  mutate(
    predicted_home_win = predicted_margin > 0,
    actual_home_win = actual_margin > 0,
    correct_prediction = predicted_home_win == actual_home_win
  ) %>%
  group_by(season) %>%
  summarize(
    games = n(),
    correct = sum(correct_prediction),
    accuracy = mean(correct_prediction),
    .groups = "drop"
  )

srs_accuracy
```



## Game Control Training Model

```{r}
# Load college football play-by-play data
pbp_raw <- load_cfb_pbp(seasons = 2021:2025)

# Filter and prepare play-by-play data
pbp_clean <- pbp_raw %>%
  filter(
    season_type == "regular",
    period %in% 1:4,
    home_team_division == "fbs",
    away_team_division == "fbs",
    !is.na(clock_minutes),
    !is.na(clock_seconds),
    clock_minutes >= 0,
    clock_minutes <= 15,
    clock_seconds >= 0,
    clock_seconds <= 59,
    !is.na(score_diff),
    !is.na(pos_team),
    !is.na(home_team),
    !is.na(away_team)
  ) %>%
  mutate(
    game_seconds_remaining =
      (4 - period) * 900 +
      (clock_minutes * 60) +
      clock_seconds,
    
    home_lead = ifelse(
      pos_team == home_team,
      score_diff,
      -score_diff
    )
  ) %>%
  arrange(season, game_id, desc(game_seconds_remaining)) %>%
  group_by(season, game_id) %>%
  mutate(
    play_duration =
      lag(game_seconds_remaining, default = 3600) -
      game_seconds_remaining,
    
    play_duration = ifelse(
      play_duration < 0,
      0,
      play_duration
    )
  ) %>%
  ungroup()

# Calculate Game Control for each game
game_control_by_game <- pbp_clean %>%
  group_by(
    season,
    week,
    game_id,
    home_team,
    away_team
  ) %>%
  summarize(
    total_time = sum(play_duration, na.rm = TRUE),
    home_gc =
      sum(home_lead * play_duration, na.rm = TRUE) /
      ifelse(total_time > 0, total_time, 1),
    away_gc = -home_gc,
    .groups = "drop"
  ) %>%
  filter(total_time >= 3000)
```

```{r}
gc_train_games <- game_control_by_game %>%
  filter(week <= 10)

gc_test_games <- game_control_by_game %>%
  filter(week > 10)
```

```{r}
calc_gc_train <- function(df) {
  teams <- unique(c(df$home_team, df$away_team))
  n_games <- nrow(df)
  
  # Construct design matrix: Home (+1), Away (-1)
  X <- matrix(0, nrow = n_games, ncol = length(teams))
  colnames(X) <- teams
  
  for (i in seq_len(n_games)) {
    X[i, df$home_team[i]] <- 1
    X[i, df$away_team[i]] <- -1
  }
  
  # Home Game Control is the response
  fit <- lm(df$home_gc ~ 1 + X)
  
  coefs <- coef(fit)
  
  home_effect <- unname(coefs["(Intercept)"])
  
  gc_vec <- coefs[-1]
  names(gc_vec) <- gsub("^X", "", names(gc_vec))
  gc_vec[is.na(gc_vec)] <- 0
  
  gc_centered <- gc_vec - mean(gc_vec, na.rm = TRUE)
  
  data.frame(
    team = names(gc_centered),
    season = df$season[1],
    Adj_GC = unname(gc_centered),
    home_effect = home_effect,
    stringsAsFactors = FALSE
  )
}
```

```{r}
gc_train_ratings <- gc_train_games %>%
  group_split(season) %>%
  map(calc_gc_train) %>%
  bind_rows()
```


### Game Control Test Predictions

```{r}
gc_test_predictions <- gc_test_games %>%
  left_join(
    gc_train_ratings %>%
      select(
        season,
        home_team = team,
        home_gc_rating = Adj_GC,
        home_effect
      ),
    by = c("season", "home_team")
  ) %>%
  left_join(
    gc_train_ratings %>%
      select(
        season,
        away_team = team,
        away_gc_rating = Adj_GC
      ),
    by = c("season", "away_team")
  ) %>%
  mutate(
    predicted_gc = home_effect + home_gc_rating - away_gc_rating,
    predicted_home_win = predicted_gc > 0
  )
```

### Game Control Out-of-Sample Accuracy

```{r}
gc_accuracy_data <- gc_test_predictions %>%
  left_join(
    test_games %>%
      select(
        season,
        week,
        home_team,
        away_team,
        home_points,
        away_points
      ),
    by = c(
      "season",
      "week",
      "home_team",
      "away_team"
    )
  ) %>%
  mutate(
    actual_home_win = home_points > away_points,
    correct_prediction = predicted_home_win == actual_home_win
  )

gc_accuracy <- gc_accuracy_data %>%
  group_by(season) %>%
  summarize(
    games = n(),
    correct = sum(correct_prediction),
    accuracy = mean(correct_prediction),
    .groups = "drop"
  )

gc_accuracy
```



## AWP Training Model

```{r}
# Calculate average win probability for each game
awp_by_game <- pbp_raw %>%
  filter(
    season_type == "regular",
    period %in% 1:4,
    home_team_division == "fbs",
    away_team_division == "fbs",
    !is.na(home_wp_before),
    !is.na(away_wp_before),
    !is.na(home_team),
    !is.na(away_team)
  ) %>%
  group_by(
    season,
    week,
    game_id,
    home_team,
    away_team
  ) %>%
  summarize(
    home_awp = mean(home_wp_before, na.rm = TRUE),
    away_awp = mean(away_wp_before, na.rm = TRUE),
    .groups = "drop"
  )

awp_train_games <- awp_by_game %>%
  filter(week <= 10)

awp_test_games <- awp_by_game %>%
  filter(week > 10)
```

```{r}
calc_awp_train <- function(df) {
  teams <- unique(c(df$home_team, df$away_team))
  n_games <- nrow(df)
  
  # Construct design matrix: Home (+1), Away (-1)
  X <- matrix(0, nrow = n_games, ncol = length(teams))
  colnames(X) <- teams
  
  for (i in seq_len(n_games)) {
    X[i, df$home_team[i]] <- 1
    X[i, df$away_team[i]] <- -1
  }
  
  # Home team's average win probability is the response
  fit <- lm(df$home_awp ~ 1 + X)
  
  coefs <- coef(fit)
  
  home_effect <- unname(coefs["(Intercept)"])
  
  awp_vec <- coefs[-1]
  names(awp_vec) <- gsub("^X", "", names(awp_vec))
  awp_vec[is.na(awp_vec)] <- 0
  
  awp_centered <- awp_vec - mean(awp_vec, na.rm = TRUE)
  
  data.frame(
    team = names(awp_centered),
    season = df$season[1],
    Adj_AWP = unname(awp_centered),
    home_effect = home_effect,
    stringsAsFactors = FALSE
  )
}
```

```{r}
awp_train_ratings <- awp_train_games %>%
  group_split(season) %>%
  map(calc_awp_train) %>%
  bind_rows()
```

### AWP Test Predictions

```{r}
awp_test_predictions <- awp_test_games %>%
  left_join(
    awp_train_ratings %>%
      select(
        season,
        home_team = team,
        home_awp_rating = Adj_AWP,
        home_effect
      ),
    by = c("season", "home_team")
  ) %>%
  left_join(
    awp_train_ratings %>%
      select(
        season,
        away_team = team,
        away_awp_rating = Adj_AWP
      ),
    by = c("season", "away_team")
  ) %>%
  mutate(
    predicted_awp = home_effect + home_awp_rating - away_awp_rating,
    predicted_home_win = predicted_awp > 0.5
  )
```

### AWP Out-of-Sample Accuracy

```{r}
awp_accuracy_data <- awp_test_predictions %>%
  left_join(
    test_games %>%
      select(
        season,
        week,
        home_team,
        away_team,
        home_points,
        away_points
      ),
    by = c(
      "season",
      "week",
      "home_team",
      "away_team"
    )
  ) %>%
  mutate(
    actual_home_win = home_points > away_points,
    correct_prediction = predicted_home_win == actual_home_win
  )

awp_accuracy <- awp_accuracy_data %>%
  group_by(season) %>%
  summarize(
    games = n(),
    correct = sum(correct_prediction),
    accuracy = mean(correct_prediction),
    .groups = "drop"
  )

awp_accuracy
```


## Model Comparison

```{r}
model_comparison <- tibble(
  Model = c(
    "SRS",
    "Adjusted Game Control",
    "Adjusted AWP"
  ),
  Accuracy = c(
    mean(
      (srs_test_predictions$predicted_margin > 0) ==
      (srs_test_predictions$actual_margin > 0)
    ),
    mean(gc_accuracy_data$correct_prediction),
    mean(awp_accuracy_data$correct_prediction)
  )
) %>%
  mutate(
    Accuracy = round(Accuracy, 3)
  ) %>%
  arrange(desc(Accuracy))

kable(
  model_comparison,
  format = "markdown",
  col.names = c("Model", "Accuracy"),
  align = c("l", "r")
)
```


### Conclusion

I found that the SRS model had the best out-of-sample performance, with an  accuracy of  72.4%. The adjusted Game Control model was a close second at about 71.9% accuracy, and the adjusted AWP model followed with an accuracy of about 70.7%.

I chose to train the models on Weeks 1 through 10 of each season and test them on the remaining regular-season games. I chose this split because it gives each model enough early-season games to build team ratings before testing how well those ratings predict later games. It also keeps the training games before the test games, which is more realistic than randomly mixing games from the entire season.
