---
title: "Homework 4 - Question 4"
author: "Zach Perkins"
format:
  html:
    toc: true
    theme: cosmo
execute:
  output: false
  warning: false
  message: false
---

# Question 4 - Adjusted vs. Unadjusted Ratings

## Data Import & Preprocessing

```{r}
# Install required packages if missing:
# install.packages(c("cfbfastR", "dplyr", "tidyr", "purrr", "knitr"))

library(cfbfastR)
library(dplyr)
library(tidyr)
library(purrr)
library(knitr)

seasons <- 2021:2025
```

## Game Control

```{r}
pbp_raw <- load_cfb_pbp(seasons = 2021:2025)

pbp_clean <- pbp_raw %>%
  filter(
    season_type == "regular",
    period %in% 1:4,
    home_team_division == "fbs",
    away_team_division == "fbs",
    !is.na(clock_minutes),
    !is.na(clock_seconds),
    clock_minutes >= 0,
    clock_minutes <= 15,
    clock_seconds >= 0,
    clock_seconds <= 59,
    !is.na(score_diff),
    !is.na(pos_team),
    !is.na(home_team),
    !is.na(away_team)
  ) %>%
  mutate(
    game_seconds_remaining =
      (4 - period) * 900 +
      (clock_minutes * 60) +
      clock_seconds,
    
    home_lead = ifelse(
      pos_team == home_team,
      score_diff,
      -score_diff
    )
  ) %>%
  arrange(season, game_id, desc(game_seconds_remaining)) %>%
  group_by(season, game_id) %>%
  mutate(
    play_duration =
      lag(game_seconds_remaining, default = 3600) -
      game_seconds_remaining,
    
    play_duration = ifelse(
      play_duration < 0,
      0,
      play_duration
    )
  ) %>%
  ungroup()

game_control_by_game <- pbp_clean %>%
  group_by(
    season,
    game_id,
    home_team,
    away_team
  ) %>%
  summarize(
    total_time = sum(play_duration, na.rm = TRUE),
    home_gc =
      sum(home_lead * play_duration, na.rm = TRUE) /
      ifelse(total_time > 0, total_time, 1),
    away_gc = -home_gc,
    .groups = "drop"
  ) %>%
  filter(total_time >= 3000)
```


### Unadjusted Game Control

```{r}
gc_team_games <- bind_rows(
  game_control_by_game %>%
    select(
      season,
      game_id,
      team = home_team,
      game_control = home_gc
    ),
  
  game_control_by_game %>%
    select(
      season,
      game_id,
      team = away_team,
      game_control = away_gc
    )
)

season_game_control <- gc_team_games %>%
  group_by(season, team) %>%
  summarize(
    avg_game_control = mean(game_control, na.rm = TRUE),
    .groups = "drop"
  )
```

### Adjusted Game Control

```{r}
calc_adj_gc <- function(df) {
  teams <- unique(c(df$home_team, df$away_team))
  n_games <- nrow(df)
  
  X <- matrix(0, nrow = n_games, ncol = length(teams))
  colnames(X) <- teams
  
  for (i in seq_len(n_games)) {
    X[i, df$home_team[i]] <- 1
    X[i, df$away_team[i]] <- -1
  }
  
  fit <- lm(df$home_gc ~ 1 + X)
  
  coefs <- coef(fit)
  
  gc_vec <- coefs[-1]
  names(gc_vec) <- gsub("^X", "", names(gc_vec))
  gc_vec[is.na(gc_vec)] <- 0
  
  gc_centered <- gc_vec - mean(gc_vec, na.rm = TRUE)
  
  data.frame(
    team = names(gc_centered),
    season = df$season[1],
    Adj_GC = unname(gc_centered),
    stringsAsFactors = FALSE
  )
}

adj_gc_by_season <- game_control_by_game %>%
  group_split(season) %>%
  map(calc_adj_gc) %>%
  bind_rows()
```


### Game Control Correlations

```{r}
gc_comparison <- season_game_control %>%
  inner_join(
    adj_gc_by_season,
    by = c("season", "team")
  )

gc_correlation_by_season <- gc_comparison %>%
  group_by(season) %>%
  summarize(
    correlation = cor(
      avg_game_control,
      Adj_GC,
      use = "complete.obs"
    ),
    .groups = "drop"
  )

gc_correlation_all <- cor(
  gc_comparison$avg_game_control,
  gc_comparison$Adj_GC,
  use = "complete.obs"
)

gc_correlation_by_season
gc_correlation_all
```


## Average Win Probability

```{r}
# Calculate game-level average win probability
awp_by_game <- pbp_raw %>%
  filter(
    season_type == "regular",
    period %in% 1:4,
    home_team_division == "fbs",
    away_team_division == "fbs",
    !is.na(home_wp_before),
    !is.na(away_wp_before),
    !is.na(home_team),
    !is.na(away_team)
  ) %>%
  group_by(
    season,
    game_id,
    home_team,
    away_team
  ) %>%
  summarize(
    home_awp = mean(home_wp_before, na.rm = TRUE),
    away_awp = mean(away_wp_before, na.rm = TRUE),
    .groups = "drop"
  )

# Reshape into one row per team per game
awp_team_games <- bind_rows(
  awp_by_game %>%
    select(
      season,
      game_id,
      team = home_team,
      awp = home_awp
    ),
  
  awp_by_game %>%
    select(
      season,
      game_id,
      team = away_team,
      awp = away_awp
    )
)

# Unadjusted season AWP
season_awp <- awp_team_games %>%
  group_by(season, team) %>%
  summarize(
    avg_wp = mean(awp, na.rm = TRUE),
    .groups = "drop"
  )
```

### Adjusted AWP

```{r}
calc_adj_awp <- function(df) {
  teams <- unique(c(df$home_team, df$away_team))
  n_games <- nrow(df)
  
  X <- matrix(0, nrow = n_games, ncol = length(teams))
  colnames(X) <- teams
  
  for (i in seq_len(n_games)) {
    X[i, df$home_team[i]] <- 1
    X[i, df$away_team[i]] <- -1
  }
  
  fit <- lm(df$home_awp ~ 1 + X)
  
  coefs <- coef(fit)
  
  awp_vec <- coefs[-1]
  names(awp_vec) <- gsub("^X", "", names(awp_vec))
  awp_vec[is.na(awp_vec)] <- 0
  
  awp_centered <- awp_vec - mean(awp_vec, na.rm = TRUE)
  
  data.frame(
    team = names(awp_centered),
    season = df$season[1],
    Adj_AWP = unname(awp_centered),
    stringsAsFactors = FALSE
  )
}

adj_awp_by_season <- awp_by_game %>%
  group_split(season) %>%
  map(calc_adj_awp) %>%
  bind_rows()
```



### AWP Correlations

```{r}
awp_comparison <- season_awp %>%
  inner_join(
    adj_awp_by_season,
    by = c("season", "team")
  )

awp_correlation_by_season <- awp_comparison %>%
  group_by(season) %>%
  summarize(
    correlation = cor(
      avg_wp,
      Adj_AWP,
      use = "complete.obs"
    ),
    .groups = "drop"
  )

awp_correlation_all <- cor(
  awp_comparison$avg_wp,
  awp_comparison$Adj_AWP,
  use = "complete.obs"
)

awp_correlation_by_season
awp_correlation_all
```



## Correlation Comparison

```{r}
correlation_comparison <- gc_correlation_by_season %>%
  rename(
    `Game Control` = correlation
  ) %>%
  left_join(
    awp_correlation_by_season %>%
      rename(
        AWP = correlation
      ),
    by = "season"
  ) %>%
  mutate(
    `Game Control` = round(`Game Control`, 3),
    AWP = round(AWP, 3)
  )

all_seasons_row <- tibble(
  season = "All",
  `Game Control` = round(gc_correlation_all, 3),
  AWP = round(awp_correlation_all, 3)
)

correlation_table <- bind_rows(
  correlation_comparison %>%
    mutate(season = as.character(season)),
  all_seasons_row
)

kable(
  correlation_table,
  format = "markdown",
  col.names = c(
    "Season",
    "Game Control Correlation",
    "AWP Correlation"
  ),
  align = c("c", "r", "r")
)
```


### Conclusion

I found that both adjusted models were strongly correlated with their unadjusted versions. For all of the seasons combined, the correlation was about 0.866 for Game Control and 0.874 for AWP. Also, the correlations were high in each individual season from 2021 to 2025.

This means that adjusting for opponent strength changed the ratings some, but the overall results stayed somewhat similar. Teams that were rated highly before the adjustment were usually still rated highly after the adjustment. Overall, the AWP model had a slightly higher correlation than Game Control.